Bounds, Semantics/Pragmatics-Divide, and Evolution

### 2016-08-02 paper/paper.tex
updated introduction draft for paper

In linguistic theorizing, it is common to draw a distinction between semantics and pragmatics. Broadly speaking, the former concerns the truth-conditional content of expressions, whereas the latter concerns information beyond such literal meanings. For instance, the integration of context and the derivation of defeasible inferences. Thus, the information conveyed by an utterance is seldom, if ever, solely determined by semantics, but rather by its pragmatic enrichment. For instance, truth-conditionally, {\em Bill read five books} does not set an upper-bound to the amount of books that Bill read; he may have read six, seven, or more books. However, the (defeasible) inference that {\em Bill read no more than five books} may be drawn on pragmatic grounds. After all, if Bill had read, for example, six books, and had the speaker known this, she would have said so. In this manner, what is conveyed (pragmatically) can go beyond what is (literally) said \citep{grice:1975}.

@@

Prima facie, it is puzzling that {\em Bill read five books} does not semantically rule out its more informative or ``stronger'' alternatives. More pointedly, would it not serve language users better if weak(er) expressions such as {\em warm}, {\em or}, {\em some} or {\em big} were not truth-conditionally compatible with stronger alternatives such as, respectively, {\em hot}, {\em and}, {\em all} and {\em huge}? After all, analogously to the example sketched above, this is what a pragmatically enriched interpretation yields. Once a distinction between semantics and pragmatics is drawn, similar questions can be posed for other types of pragmatic inferences, such as ignorance inferences as (\ref{ex:ignorance}), or manner inferences such as (\ref{ex:manner-1}) and (\ref{ex:manner-2}).

In more general terms, the question then is what motivates the selection of particular semantic structures over other hypothetical alternatives in light of pragmatics. While a number of recent investigations have began to address issues pertaining to the development and selection of linguistic features, the semantics-pragmatics distinction has received little attention (see \citealt{steels:2015} and \citealt{tamariz+kirby:2016} for recent overviews). Our starting point is given by the overarching account of competing pressures that crystallizes across different frameworks: Natural languages need to allow for successful communication, one the one hand, and be well suited for acquisition on the other. Put differently, languages need to be well-adapted to the communicative needs within a linguistic community, and be learnable to survive their faithful transmission across generations.

 \hl{Description of plan to lay out model and test it on upper-bounds.}

\hl{general notes on cultural evolution, biological evolution, and their connection}

@@

\subsection{Probabilistic models of language use}

>>>>

In linguistic theorizing, it is common to draw a distinction between semantics and pragmatics. Broadly speaking, the former concerns the truth-conditional content of expressions, whereas the latter concerns information beyond such literal meanings. For instance, the integration of context and the derivation of defeasible inferences. Thus, the information conveyed by an utterance is seldom, if ever, solely determined by semantics, but rather by its pragmatic enrichment. For instance, truth-conditionally, {\em Bill read five books} does not set an upper-bound to the amount of books that Bill read; he may have read six, seven, or more books. However, the (defeasible) inference that {\em Bill read no more than five books} may be drawn on pragmatic grounds. After all, if Bill had read, for example, six books, and had the speaker known this, she would have said so. In this manner, what is conveyed (pragmatically) can go beyond what is (literally) said.

@@

Prima facie, it is puzzling that {\em Bill read five books} does not semantically rule out its more informative or ``stronger'' alternatives. More poignantly, would it not serve language users better if weak(er) expressions such as {\em warm}, {\em or}, {\em some} or {\em big} were truth-conditionally incompatible with stronger alternatives such as, respectively, {\em hot}, {\em and}, {\em all} and {\em huge}? After all, analogously to the example sketched above, this is what a pragmatically enriched interpretation yields; an upper-bound that rules out these alternatives through a so-called scalar inference \citep{horn:1972,gazdar:1979}. Once a distinction between semantics and pragmatics is drawn, similar questions can be posed for other types of pragmatic inferences, such as ignorance inferences as and manner inferences.

A number of recent investigations have began to address issues pertaining to the development and selection of linguistic properties (see \citealt{steels:2015} and \citealt{tamariz+kirby:2016} for recent overviews). Our starting point is given by the overarching account of competing pressures that has crystallized across these approaches: Natural languages need to allow for successful communication, one the one hand, and be well suited for acquisition, on the other. That is, they need to be well-adapted to the communicative needs within a linguistic community, as well as be learnable to survive their faithful transmission across generations. We proceed by instantiating these pressures in the replicator-mutator dynamics, commonly used in evolutionary game theory. Overall, this model allows for general and precise means to model the dynamics of linguistic pressures by combining functional pressure on successful communication, effects of learning biases on (iterated) Bayesian learning \citep{griffiths+kalish:2007}, and  probabilistic models of language use in populations with distinct lexica \citep{frank+goodman:2012,franke+jaeger:2014, bergen+etal:2016}. Drawing from the latter research strand allows for the representation of a distinction between semantics and pragmatics in communicative behavior. More broadly, it connects the recent surge in models of synchronic probabilistic rational language use with diachronic models of cultural evolution.

\section{Conveying upper-bounds}
\hl{start bottom up -- from the behavior of agents to the overall dynamics. Here, begin with language use and introduce scalar inferences on the way as the object of study we will focus on here}

@@
